---
title: "PSY6422 Global Smoking Prevalence Analysis"
date: "2024-12-06"
output: html_document
flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
---

Column {data-width=250}
-----------------------------------------------------------------------

### Navigation

- [Background Information](#background-information)
- [Research Question](#research-question)
- [Project Organisation](#project-organisation)
- Dataset
  - [Dataset Origins](#dataset-origins)
  - [Limitations](#limitations)
- Data Preparation
  - [Installing Packages](#installing-packages)
  - [Libraries](#libraries)
  - [Load Data](#load-data)
  - [Sanity Check](#sanity-check)
  - [Clean Data](#clean-data)
  - [Merge Data](#merge-data)
  - [Description of Data](#description-of-data)
- [Initial Visualization](#initial-visualisation)
- [Final Visualization](#final-visualisation)

Column {data-width=750}
-----------------------------------------------------------------------
### Background Information {#background-information}

explain why I chose smoking data and maybe where my data comes from??
This data package contains the data that powers the chart ["Share of adults who smoke"](https://ourworldindata.org/grapher/share-of-adults-who-smoke?v=1&csvType=full&useColumnShortNames=false) on the Our World in Data website
https://datacatalog.worldbank.org/search/dataset/0037712/World-Development-Indicators  

---

### Research Question {#research-question}

How does the prevalence of smoking in adults across the world vary in the dataset, and what trends or patterns emerge from this analysis? 

---

### Project Organisation {#project-organisation}

The project repository includes:
- **Codebook:** Detailed variable descriptions.
- **Data Folder:** Contains the raw and processed datasets.
- **Figures Folder:** Visualizations created during analysis.
- **Scripts Folder:** R scripts for cleaning, analysis, and plotting.


The PSY6422_smoke repository is organized into key sections to help you navigate its contents. The /codebook folder provides detailed documentation on the dataset, including variable descriptions and structure, offering essential context for the analysis. The /data folder contains the raw datasets used in this project, forming the basis of all analyses. The /figures folder showcases visualizations and plots created during the analysis, highlighting the project’s key findings and insights. Lastly, the /scripts folder includes all the code used for data processing, analysis, and visualization. Together, these sections guide you through the project workflow, from raw data to final outputs.

---

### Dataset

#### Dataset Origins {#dataset-origins}

The raw dataset for this visualization project comes from : Multiple sources compiled by World Bank (2024) – processed by Our World in Data. “Prevalence of current tobacco use (% of adults)” [dataset]. World Health Organization (via World Bank), “World Development Indicators” [original data].
Source: Multiple sources compiled by World Bank (2024) – processed by Our World In Data

The percentage of the population ages 15 years and over who currently use any tobacco product (smoked and/or smokeless tobacco) on a daily or non-daily basis. Tobacco products include cigarettes, pipes, cigars, cigarillos, waterpipes (hookah, shisha), bidis, kretek, heated tobacco products, and all forms of smokeless (oral and nasal) tobacco. Tobacco products exclude e-cigarettes (which do not contain tobacco), “e-cigars”, “e-hookahs”, JUUL and “e-pipes”. The rates are age-standardized to the WHO Standard Population.

#### Limitations {#limitations}

- Large uncertainty in countries with irregular surveys or missing data.
- Estimates exclude e-cigarettes and other non-tobacco products.

These considerations are important when interpreting the project's results 
Estimates for countries with irregular surveys or many data gaps have large uncertainty ranges, and such results should be interpreted with caution.

---

### Data Preparation

#### Installing Packages {#installing-packages}

```{r install and load packages}
# List of packages to install and load
packages <- c("tidyverse", "ggplot2", "tidyr", "dplyr", "plotly", "rnaturalearth", "rnaturalearthdata", "sf", "htmlwidgets")

# Function to install packages and load them
install_and_load <- function(packages) {
  for (package in packages) {
    if (!require(package, character.only = TRUE)) {
      install.packages(package, dependencies = TRUE)
      library(package, character.only = TRUE)
    } else {
      library(package, character.only = TRUE)
    }
  }
}

# Run the function
install_and_load(packages)
```
``` 

---

### Libraries {#libraries} 

---

### Load Data {#load-data}

Code to load the dataset and a brief description of the file used.

```{r load the data}
# Load raw data
rawdata <- read.csv("data/smoking.csv")
```

---

### Sanity Check {#sanity-check}

Summary and structure of the dataset before cleaning.

```{r sanity check}
#sanity check 
str(rawdata)           # Inspect structure and summary
summary(rawdata)
head(rawdata)          # Check the first few rows
dim(rawdata)           # Check dimensions
colSums(is.na(rawdata))# Check for missing values

# Check for duplicates
duplicates <- rawdata[duplicated(rawdata), ]
print(duplicates)

# Unique values in key columns
unique(rawdata$Entity)
unique(rawdata$Year)
```
---

### Clean Data {#clean-data}

Steps and code for cleaning the data.
```{r cleaning data}
# Cleaning the data 
# Remove specific entities
countries_data <- rawdata[!rawdata$Entity %in% c("East Asia and Pacific (WB)", "Sub-Saharan Africa (WB)", 
                                                 "Upper-middle-income countries", "Europe and Central Asia (WB)", 
                                                 "World", "European Union (27)", "Low-income countries", 
                                                 "Lower-middle-income countries", "Middle East and North Africa (WB)", 
                                                 "Middle-income countries", "North America (WB)", "South Asia (WB)", 
                                                 "Latin America and Caribbean (WB)", "High-income countries"), ]

# Exclude specific years
countries_data <- countries_data %>%
  filter(!(Year %in% c(2018, 2019)))

# Rename column for ease of use
countries_data <- countries_data %>%
  rename(Prevalence = Prevalence.of.current.tobacco.use....of.adults.)

```

# Loading and cleaning world data in order to merge with smoke data
```{r}
# Load world map data
world <- ne_countries(scale = "medium", returnclass = "sf")

# Identify codes in countries_data but not in world
missing_in_world <- setdiff(countries_data$Code, world$iso_a3)
print(missing_in_world)

# Identify codes in world but not in countries_data
missing_in_data <- setdiff(world$iso_a3, countries_data$Code)
print(missing_in_data)

# Fix missing ISO codes
fix_iso_codes <- function(world_data) {
  world_data %>%
    mutate(iso_a3 = ifelse(name == "France", "FRA", iso_a3)) %>%
    mutate(iso_a3 = ifelse(name == "Norway", "NOR", iso_a3))
}
world <- fix_iso_codes(world)
```

---

### Merge Data {#merge-data}

Explanation and code for merging the datasets.
```{r}
# Merge datasets, sanity check/clean merged data 
# Merge datasets
map_data <- world %>%
  left_join(countries_data, by = c("iso_a3" = "Code"))

# Replace NA prevalence values with 0
map_data <- map_data %>%
  mutate(Prevalence = ifelse(is.na(Prevalence), 0, Prevalence))

# Remove duplicates
map_data <- map_data[!duplicated(map_data), ]

# Remove geometry for Plotly compatibility
plot_data <- map_data %>%
  st_set_geometry(NULL)

# Structure and summary of the cleaned dataset
str(plot_data)
summary(plot_data)

# Check for duplicates in country-year pairs
duplicates <- plot_data %>%
  group_by(iso_a3, Year) %>%
  filter(n() > 1)
print(duplicates)

# Check for missing prevalence values
missing_prevalence <- plot_data %>%
  filter(is.na(Prevalence))
print(missing_prevalence)

# Distribution of years and countries
table(plot_data$Year)
table(plot_data$iso_a3)

```

---

### Description of Data {#description-of-data}

Summary of the cleaned data, including key variables.

# creating variables 

In order to create my visualization, I had to create new variables. 
I got variable world from the function sf, which contains the resources necessary for my analysis. 
I then merged my world data and my smoking data into 'map_data' via the ISO code. 


# cleaning data
after my initial sanity check, I started to clean my data. 
The data contained specific entities such as different regions of 
the globe and the different income levels. In order to visualize the data I had to take these out. Further more, 
I wanted to visualize the data by 5 years so i got rid of 2018 and 2019.
I also renamed the variable prevalence for ease.and fixed my missing isos


# description of the cleaned data 
Entity refers to the name of the country. Code refers to the OWID internal entity code that we use if the entity is a country or region. Year refers to the years of the prevalence. Prevalence.of.current.tobacco.use....of.adults. refers to the prevalence of current tobacco users. 

---

### Initial Visualisation {#initial-visualisation}

Code and output for the initial visualisation.

### Visualization {#visualization}
This is the very first visualization I got for my data. 
The initial visualization provided a basic representation of smoking prevalence on a static map. While it successfully displayed the prevalence rates for countries, it lacked interactivity and failed to account for changes across years. Users could not explore how smoking prevalence evolved over time or access specific data points for each country. The map presented a single snapshot without the ability to delve into detailed information, such as individual country names or exact prevalence percentages. These limitations highlighted the need for a more dynamic and informative visualization to enhance user engagement and understanding.

```{r}
ggplot(data = map_data) +
  geom_sf(aes(fill = Prevalence), color = "white", size = 0.2) +
  scale_fill_gradient(low = "blue", high = "red", na.value = "grey") +
  labs(title = "Global Smoking Prevalence Heatmap",
       fill = "Prevalence (%)") +
  theme_minimal()

```

---

### Final Visualisation {#final-visualisation}
## Final visualization 


